Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-10 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
small edits

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. %clarify share
Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

\section{A Verified Binary Search Tree with Hand-over-Hand Locking}
\label{hand-over-hand}

We demonstrate our approach by verifying a binary search tree (BST) with fine-grained locking in VST, using the above approach to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

@@

Now we have everything we need to specify the tree operations:
\begin{mathpar}
\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle

\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle

\forall m.\,\langle \nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
\end{mathpar}
where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g$$
Now we are prepared to prove that each of our BST functions satisfies its specification.

@@

\subsubsection{Insert}
An outline of the C code for the insert function is shown in figure \ref{bst-conc}. When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking. If it encounters a node with key $k$, then it simply changes the value at that node. Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node. The atomic specification for insert method is $$\forall m.\ \langle \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.

@@

The code for the $\mathsf{lookup}$ method is shown in Figure
\ref{lookupproof}. It takes the location of the root pointer and a key
as the arguments. A thread spans the tree to find a given key using
hand-over-hand locking mechanism. Once a thread finds the key in the
tree, it gets the value associated with that key, releases the current
node's lock, and returns the value.
The atomic specification for $\mathsf{lookup}$ is
$$\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle$$
The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.

@@

The atomic specification for \lstinline{delete} is:
$$\forall t.\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$

>>>>

\subsection{Logically Atomic Specifications}
\label{atomic}

%revise
For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. %describe weird binder
Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
\begin{mathpar}
\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure. %a bit more about how we prove these, and what the next couple sections are doing about them

\subsection{Atomicity with lock invariants}%make clearer the connection to atomic specs
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. %clarify share
Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

\subsection{Atomicity with atomic locks}
Proving atomic specifications for a data structure using atomic specifications for locks is more immediate. As an example, consider Krishna et al.'s verification of a hand-over-hand locking pattern for search structures~\cite{templates}. The lock specification used in the paper is $$\langle \mathsf{inFP}(n)\ |\ \mathsf{CSS}(r, C) \rangle\ \mathtt{lockNode}\ n\ \langle \mathsf{CSS}(r, C) * \mathsf{N}(n)\rangle$$ where $\mathsf{inFP}(n)$ asserts that the node $n$ is in the data structure $\mathsf{CSS}$ is the abstract state, and $\mathsf{N}(n)$ is the resources for the node $n$ (including both points-to assertions and ghost state). The abstract state assertion $\mathsf{CSS}$ includes both global ghost state and the composition of per-node state $\mathsf{N}$ for each node that is not currently locked, so that $\mathsf{N}(n)$ can be removed from the abstract state when $n$'s lock is acquired.

@@

\section{A Verified Binary Search Tree with Hand-over-Hand Locking}
\label{hand-over-hand}%show code, somewhere, maybe cut-down pseudocode

We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

%make this more concise, and maybe merge it into a later section
This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n'$'s %awkward
 lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

@@

Now we have everything we need to specify the tree operations:
\begin{mathpar}
\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{insert}\ m\ k\ v)\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep_g\,p\ |\ \treerep_g\,m \land \mathrm{lookup}\,m\ k = v\rangle

\langle m.\,\nodeboxrep_g\,p\ |\ \treerep_g\,m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\,p\ |\ \treerep_g\,(\mathrm{delete}\ m\ k)\rangle
\end{mathpar}
where $\treerep_g\,m$ asserts that global ghost state of the tree implements a map $m$, as described above, and $\nodeboxrep$ (the client's handle to the data structure) is simply a reference to the lock for the root node:
$$\nodeboxrep_g\,b \triangleq \exists p.\ b \mapsto p * p.\texttt{lock} \islock \mathsf{node\_inv}_g$$
Now we are prepared to prove that each of our BST functions satisfies its specification.

@@

\subsubsection{Insert}
An outline of the C code for the insert function is shown in figure \ref{bst-conc}. When a thread inserts a key $k$ into a tree, it first crawls the tree to find the right position for $k$ using hand-over-hand locking. If it encounters a node with key $k$, then it simply changes the value at that node. Otherwise, it will eventually reach a leaf node, where it creates two new empty leaf nodes, inserts the key and value into the current node, and links the newly created leaf nodes to the current node. The atomic specification for insert method is $$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ m\ k\ v)\rangle$$
The \lstinline{insert} function should atomically add the key-value pair $(k, v)$ to the map implemented by the BST at $p$.

@@

The code for the $\mathsf{lookup}$ method is shown in Figure
\ref{lookupproof}. It takes the location of the root pointer and a key
as the arguments. A thread spans the tree to find a given key using
hand-over-hand locking mechanism. Once a thread finds the key in the
tree, it gets the value associated with that key, releases the current
node's lock, and returns the value.
The atomic specification for $\mathsf{lookup}$ is
$$\langle m.\,\nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{lookup}\ m\ k = v\rangle$$
The \texttt{lookup} function should atomically look up the value of the key $k$ in the tree at $p$.

@@

The atomic specification for \lstinline{delete} is:
$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$
